---
title: "Interactive Tornado Map of the United States"
author: "hangulandy"
date: "6/30/2017"
output: 
        html_document:
                fig_width: 8.5
                fig_height: 6
                toc: true
                number_sections: true
                toc_float:
                        collapsed: true
                        smooth_scroll: true
                keep_md: true
---

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      cache = TRUE)
source("project1.R") 
```

# Executive Summary

This project creates an interactive map showing tornado events in the United States.

All code for non-standard functions used in this anaylsis can be found in **Appendix 1**.

# Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]
There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Load and Tidy the Data

```{r}
load_packages()
dat <- load_data()

df <- add_year(dat)
df$DATE <- mdy_hms(df$BGN_DATE)
df <- tidy_data(df)
df <- subset(df[df$LATITUDE > 2500 & df$LATITUDE < 5000 & df$EVCAT == "Tornado", ])
df <- calc_damages(df)

## Need to fix longitude to convert to proper pos / neg for leaflet
df$LATITUDE <- df$LATITUDE / 100
df$LONGITUDE <- (-1 * df$LONGITUDE) / 100
```

## Create Popup Vector

```{r}
content <- paste(sep = "",
                 "Date:", df$DATE, "<br/>",
                 "Fatalities: ", df$FATALITIES, "<br/>", 
                 "Injuries: ", df$INJURIES, "<br/>",
                 "Crop Dmg: ", dollar(df$Crop), "<br/>",
                 "Prop Dmg: ", dollar(df$Property), "<br/>")
```

# Map

```{r}
df %>%
        leaflet %>% 
        addTiles %>% 
        addMarkers(lat = df$LATITUDE, lng = df$LONGITUDE, popup = content, 
                   clusterOptions = markerClusterOptions())
```

# Appendix 1: Script for Functions

```{r eval=FALSE}
## These first two lines of code set the working directory to the location of 
## the file being sourced
this.dir <- dirname(parent.frame(2)$ofile)
setwd(this.dir)

## Load packages
load_packages <- function() {
        library(leaflet)
        library(lubridate)
        library(scales)
}

## Load data
load_data <- function() {
        mainDir <- getwd()
        subDir <- "data"
        
        if (!file.exists(subDir)) {
                dir.create(file.path(mainDir, subDir))
        }
        
        setwd(file.path(mainDir, subDir))
        
        U <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        dest <- "./StormData.csv.bz2"
        
        ## Only use when downloading.  If file exists, just skip this line
        if (!file.exists(dest)) {
                download.file(U, dest) 
        }

        temp <- read.csv(dest, header = TRUE, sep = ",")
        
        setwd(mainDir)
        
        temp <- temp[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", 
                         "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", 
                         "LATITUDE", "LONGITUDE", "REMARKS")]
        
        return(temp)
}

## Takes a year variable from the date and adds it to the data table
add_year <- function(dat) {
        output <- data.frame()
        temp <- mdy_hms(dat$BGN_DATE)
        temp <- year(temp)
        
        dat$YEAR <- as.factor(temp)
        
        output <- dat
        
        return(output)
}

## Tidy the data with very specific transformations
tidy_data <- function(dat) {
        
        dat <- subset(dat[dat$LATITUDE > 0 & dat$LONGITUDE > 0, ])
        ## Coerce the remarks to character.  I did this because RStudio kept 
        ## crashing when I was reading the remarks of class factor. I coerced 
        ## to character and no longer had any issues.
        dat$REMARKS <- as.character(dat$REMARKS) 
        
        ## Create a character variable to group events since I will replace the 
        ## value with another character variable as seen in subsequent code.
        dat$EVCAT <- as.character(dat$EVTYPE) 
        
        ## These search terms are out of order compared to the documentation to 
        ## give priority. Lower in order here signifies more unique, so it is 
        ## easier to use the order to recategorize
        
        ## Heat and drought group
        ## Captures all heat events
        hea <- c(".*heat.*", ".*warm.*", ".*hot.*", ".*record temp.*", 
                 "HIGH TEMP.*", "RECORD HIGH.*") 
        ## Re-categorizes drought events (some are also labeled with heat)
        drt <- c(".*droug.*", ".*dry.*", ".*driest.*", "^EXCESSIVE$") 
        
        ## Rain, wind, thunderstorm, and flood group.  
        ## Captures many events that follow, but they will all be recategorized
        swd <- c(".*wind.*", "^WND$", "SEVERE TURBULENCE") 
        ## Captures all lightning events, some will be recategorized later
        ltn <- c(".*lightn.*", ".*lighting.*", "LIGNTNING") 
        ## Captures some flood and thunderstorm events
        rai <- c(".*heavy.*rain.*", ".*torrent.*", ".*wet.*", ".*rain.*", 
                 ".*shower.*", ".*heavy.*precip.*") 
        ## Recategorizes thunderstorm wind events
        tst <- c(".*thunder.*wind.*", ".*TSTM.*", ".*thunderst.*", "wall cloud", 
                 ".*microburst.*", "Metro Storm, May 26", "^HIGH$", "DOWNBURST", 
                 ".*gustnado.*", "APACHE COUNTY") 
        ## Captures all flood events
        fld <- c(".*fl.*d.*", ".*stream", ".*river", ".*rising.*", "DAM BREAK", 
                 "DAM FAILURE", ".*urban.*", "HIGH WATER") 
        ## Recategorizes lake flood events
        lsf <- c(".*lake.*flood.*") 
        ## Recategorizes coastal flooding
        cfl <- c(".oast.*fl.*d.*", ".*eros.*n.*") 
        ## Recategorizes flash flooding
        ffl <- c(".*flash.*", "DROWNING", "ICE JAM") 
        
        ## Surf, surge, tide, current, etc.
        ## Recategorizes some coastal flooding because of surf
        hsf <- c(".*surf.*", ".*swells.*", "ROUGH SEAS", "Marine Accident", 
                 "HIGH WAVES", "HEAVY SEAS", "HIGH SEAS", "ROGUE WAVE") 
        ## Recategorizes some surf events as rip current
        rpc <- c(".*current*") 
        ## Storm surge and tides, picks up astronomical low tide
        srg <- c(".*surge*", ".*tide.*") 
        ## Recategorizes astronomical low tide
        alt <- c(".*stronomical low.*") 
        
        ## Cold, fog, freezing, sleet, and snow group
        ## Picks up cold air funnels, some tornadoes, some snow, and many frost 
        ## observations
        cld <- c(".*cold.*", ".*low temp.*", ".*wind.*chill.*", ".*cool.*", 
                 ".*hyp.*therm.*") 
        ## Captures fog events that are not freezing fog
        fog <- c("^fog$", ".*dense fog.*", "fog and cold.*") 
        ## Captures freezing fog and ice fog events
        ffg <- c("freezing fog", "ice fog") 
        frz <- c("frost", ".*freez.*", ".*glaze.*", "^BLACK ICE$", "PATCHY ICE", 
                 ".*ic.*road.*")
        ## Captures some snow-related events
        slt <- c(".*sleet.*", ".*pellet.*") 
        ## Captures some blizzard and heavy snow events
        ist <- c(".*ice.*storm.*") 
        ## Captures some avalanche and blizzard events
        snw <- c(".*heavy.*snow.*", ".*snow.*", "RECORD PRECIPITATION") 
        ## Recategorizes some snow events, captures some blizzard events
        wnt <- c(".*wint.*", ".*mix.*", "ICE FLOES", "^ICE$") 
        ## Recategorizes some snow events, captures some blizzard events
        avl <- c(".*valanc.*") 
        ## Recategorizes some snow events and all blizzard events
        blz <- c(".*lizzard.*") 
        ## Recategorizes lake effect snow events
        les <- c(".*lake.*snow.*") 
        
        ## Funnel cloud, hail, tornado group
        ## Picks up some waterspouts, thunderstorms, tornadoes and hail
        fnl <- c("funnel", ".*spout.*") 
        ## Picks up some thunderstorms, tornadoes, and flooding, and icy roads
        hai <- c(".*hail.*") 
        ## Recategorizes marine hail
        mha <- c("marine hail") 
        ## Recategorizes marine wind events of all types
        mwd <- c("marine.*wind.*", "COASTAL.*STORM", "MARINE MISHAP") 
        ## Recategorizes tornado events
        tdo <- c(".*torn.*") 
        
        ## Misc. values that are unique enough that they don't show up much
        dbr <- c(".*debris.*", ".*slide.*", ".*landsl.*")
        smk <- c(".*smoke.*")
        dst <- c(".*dust.*")
        ddv <- c(".*dust dev.*")
        sch <- c(".*seiche*")
        tsn <- c(".*tsunam.*")
        vol <- c(".*volc.*", "VOG")
        fir <- c(".*fire.*", "RED FLAG CRITERIA")
        
        ## Large storm group.  These are the highest that should overwrite 
        ## previous, lesser events
        tps <- c("tropical")
        tpd <- c(".*depress.*")
        hcn <- c(".*hurricane.*", ".*typhoon.*")
        smr <- c(".*summary.*", ".*monthly.*", "NONE", "No Severe Weather", 
                 "Other", "MILD PATTERN", ".*normal precip.*", 
                 "EXCESSIVE PRECIPITATION", "RECORD LOW", "SOUTHEAST", 
                 "NORTHERN LIGHTS")
        
        ## CANNOT c() the vectors or they will become one vector.  
        ## Must keep them separate like this so that the remain a list to use 
        ## in the loop
        searchterms <- list(swd,
                            ltn,
                            rai,
                            tst,
                            fld,
                            lsf,
                            cfl,
                            ffl,
                            hsf,
                            rpc,
                            srg,
                            alt,
                            cld,
                            fog,
                            ffg,
                            frz,
                            slt,
                            ist,
                            snw,
                            wnt,
                            avl,
                            blz,
                            les,
                            hea,
                            drt,
                            fnl,
                            hai,
                            mha,
                            mwd,
                            tdo,
                            dbr,
                            smk,
                            ddv,
                            dst,
                            sch,
                            tsn,
                            vol,
                            fir,
                            tps,
                            tpd,
                            hcn,
                            smr
        )
        
        labels <- list("Strong Wind",
                       "Lightning",
                       "Heavy Rain",
                       "T.Storm Wind",
                       "Flooding",
                       "Lakeshore Fl.",
                       "Coastal Flood",
                       "Flash Flood",
                       "High Surf",
                       "Rip Curr.",
                       "St. Surge",
                       "Astronomical Low Tide",
                       "Cold / Wind Chill",
                       "Dense Fog",
                       "Freezing Fog",
                       "Frost / Freeze",
                       "Sleet",
                       "Ice Storm",
                       "Heavy Snow",
                       "Winter Storm",
                       "Avalanche",
                       "Blizzard",
                       "Lake-Effect Snow",
                       "Heat",
                       "Drought",
                       "Funnel Cloud",
                       "Hail",
                       "Marine Hail",
                       "Marine Wind",
                       "Tornado",
                       "Debris",
                       "Dense Smoke",
                       "Dust Devil",
                       "Dust Storm",
                       "Seiche",
                       "Tsunami",
                       "Volcanic Ash",
                       "Wildfire",
                       "Trop. Storm",
                       "Trop. Dep.",
                       "Hurricane",
                       "Summaries"
        )
        
        
        for (i in 1:length(searchterms)) {
                temp <- unique(grep(paste(searchterms[i][[1]], collapse = "|"), 
                                    dat$EVTYPE, ignore.case = TRUE, 
                                    value = TRUE))
                dat$EVCAT <- replace(dat$EVCAT, dat$EVTYPE %in% temp, 
                                     as.character(labels[[i]]))
        }
        
        ## The ? entry didn't work in my grep searches, so I had to add an 
        ## extra line of code for this one observation.
        dat$EVCAT <- replace(dat$EVCAT, dat$EVTYPE == "?", 
                             as.character("Summaries")) 
        
        ## Remove all other variables from memory
        rm(list = ls()[!(ls() %in% c('dat'))])
        
        ## Reclass the EVCAT values to factor so they are usable later in other 
        ## functions
        dat$EVCAT <- as.factor(dat$EVCAT) 
        return(dat)
}


## Starting of code to recalculate damages
calc_damages <- function(dat) {
        ## Coerce from factor to character for searching so the code is easier
        dat$PROPDMGEXP <- as.character(dat$PROPDMGEXP) 
        dat$CROPDMGEXP <- as.character(dat$CROPDMGEXP)
        
        ## Create variables to show damage with default value 1000
        dat$PMult <- 1000 
        dat$CMult <- 1000
        
        ## List of all possible values for PROPDMGEXP and CROPDMGEXP that 
        ## should be multiplied by a number other than 1000
        searchterms <- list("B", "M", "m") 
        
        ## Create vector of multipliers to pair with searchterms
        multiplier <- c(1000000000, rep(1000000, 2)) 
        
        ## Replaces multiplier values with proper numbers
        for (i in 1:length(searchterms)) {
                dat$PMult <- replace(dat$PMult, dat$PROPDMGEXP == searchterms[i], multiplier[i])
        }
        
        ## Calculates crop damage based on multipliers 
        for (i in 1:length(searchterms)) {
                dat$CMult <- replace(dat$CMult, dat$CROPDMGEXP == searchterms[i], multiplier[i])
        }
        
        ## Calculate final property damage and crop damage values
        dat$Property <- dat$PROPDMG * dat$PMult 
        dat$Crop <- dat$CROPDMG * dat$CMult
        
        rm(list = ls()[!(ls() %in% c('dat'))])
        
        return(dat)
}
```


# Appendix 2: Session Information

```{r}
sessionInfo()
```

